\propertitle

### 2020-01-16 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2020-01-16T10:11:59+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{4 November 2015}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The model can be viewed as a straightforward combination of the
maximum-entropy method and basic sampling relations from the
probability-calculus, discussed in \sect~\ref{sec:model}. The
maximum-entropy or minimum-relative-entropy method
\citep{jaynes1957}[\chap~11]{jaynes1994_r2003}[for the minimum-relative
entropy method see:][\sect~4.b]{jaynes1963}[for a more rigorous
derivation]{hobsonetal1973}[for applications]{sivia1996_r2006}
%[see also][for applications to irreversible statistical mechanics]{grandy1980}
to build initial probability distributions has been used in the
neurosciences for different kinds of investigations about the neuronal
activity of various brain areas and about other phenomena of importance % ,
% for example gene and protein interaction
\citep[]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,roudietal2009c,barreiroetal2010,shimazakietal2015,moraetal2015%,lezonetal2006,weigtetal2009
}[for
further references see][]{lathametal2013}. However, our model can also be
viewed as an approximation of a more detailed model within Bayesian theory;
we shall constantly keep this point of view in mind in our discussion.

@@

% Before applying the formula above to two concrete data sets we want to add
% two remarks about the maximum-entropy method that are seldom made in the
% neuroscientific literature. They are important for the interpretation of
% the results and are further discussed in \sect~\ref{sec:assumptions}.
% First, the maximum-entropy
% method % (based on the \emph{Shannon} relative
% % entropy, as opposed to other entropy functions such as Burg's
% % \citeyear{burg1975})
% rests on some implicit assumptions about the probabilities for the long-run
% frequency distribution of activities, besides the assumptions just
% mentioned at the end of the previous paragraph. So the often-heard
% statement that it gives \enquote{the maximally unbiased (or non-committal)
%   distribution} must be taken with a grain of salt.
% \citep{jaynes1986d_r1996,portamana2009,portamana2017}. Second,
A maximum-entropy distribution like $\px(\yA \| M, N)$ is equivalent to the
zeroth-order Laplace approximation
\citep[\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four distinct distributions for the larger population, which differ
numerically from one another in higher-order approximations:
\begin{enumerate}[label=(\roman*)]
\item\label{item:prob_freq} the most probable \emph{frequency} distribution
  for the total activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item\label{item:prob_binin} the \emph{belief} distribution for the value
  of the total activity at any time bin among the \emph{recorded} ones,
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item\label{item:prob_longrun} the most probable \emph{frequency}
  distribution for the total activity in a very long run of \emph{new} time
  bins, %$\ynuu(\yA)$,
\item\label{item:prob_binout} the \emph{belief} distribution for the value
  of the total activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{enumerate}
Their mathematical relationship is presented more explicitly in
appendix~\ref{sec:inference_prob}. The
distribution~\eqref{eq:app_maxent_pop_first} can therefore be interpreted
in each of the four ways above, in this approximate sense. In the present
case the validity of the approximation decreases as the ratio $nN/T$
increases. The most robust interpretation is~\ref{item:prob_freq}, and that
is how we prefer to interpret the distribution $\px$: as \emph{the most
  probable relative-frequency distribution} of the activity $\yA$. See
\sect~\ref{sec:assumptions} for further discussion about assumptions and
approximations.

@@

Our question is whether some degree of \enquote{cooperation} is more
\enquote{important} than some other degree, or not. In terms of
informational sufficiency, its definition might suggest to compare the
belief distributions conditional on two different subsets of statistics.
But which measure should be used for such a comparison? and how to motivate
it? We prefer to take a different approach, leading to an automatic answer:
let us thus turn the informational sufficiencies of two different subsets
of statistics into two \emph{hypotheses}. We can then ask: which hypothesis
is more probable, given the data? Equivalently, what is the ratio of their
probabilities? By translating our question about \enquote{cooperativity}
into a hypothesis-testing problem about sufficiency, we can find a
well-defined answer determined by the probability-calculus. % The ratio of
% the two probabilities gives us a natural measure of the
% For example, do
% the data say that the first two empirical moments are probably enough to
% determine our beliefs? or is it more probable that the first four are
% needed instead?
We shall see that this approach is indeed more powerful than a simple
comparison of distributions.

@@

The second assumption, typical of maximum-entropy in general, concerns the
probability distribution for the long-run frequencies ensuing from the
first assumption \citep[and references therein]{portamana2017}. It is taken
to be either an \enquote{entropic prior}
\citep{skillingetal1984,rodriguez1991,neumann2007}, which takes into
account the multiplicity of long-run frequencies, or a prior for a model
with sufficient statistics, which expresses the sufficiency of the moments
under study, according to the theorem discussed in
\sect~\ref{sec:marginalization}. This assumption is crucial for the
maximum-entropy method based on the Shannon entropy. Different prior
densities lead to alternative maximum-entropy methods; for example one
based on the Burg entropy \citep{burg1975} if a Dirichlet density is used
\citep{jaynes1986d_r1996}. The entropic prior depends on a reference
distribution and on a coefficient $L$ which determines the sharpness of the
prior's peak. The computations of \sects~\ref{sec:application}
and~\ref{sec:marginalization} were insensitive to the choice of the
reference distribution in their significant digits. We used three
biologically reasonable alternatives: a flat distribution, one linearly
decreasing with $\yA$, and a normal one with a broad peak at low
activities.

@@

The two infinity approximations are at the boundary of their validity in
our examples. For this reason we interpret the maximum-entropy distribution
as the \emph{most probable frequency distributions}, as mentioned at the
end of \sect~\ref{sec:model}. An important question, then, is \emph{which
  features of the maximum-entropy frequency distribution are typical of the
  majority of plausible frequency distributions?} Especially for features
such as number, position, height of modes, or tail behaviour. Preliminary
studies using a full probability calculation (to be published separately)
show that several semi-quantitative features of the maximum-entropy
distribution are indeed typical. For the first data set, for example, most
frequency distributions have two high-frequency regimes of activity, as in
\fig~\ref{fig:plots_5mom_a}, with roughly the same heights and almost the
same locations, within a few percentages of population-averaged activity.
Each of the two high-frequency regimes, though, may consist of two or three
very close modes rather than one. The same preliminary studies also
indicate that these typical features are robust for different prior choices
in the second assumption above, for example using a Dirichlet prior instead
of an entropic one.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2020-01-16T12:46:05+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{4 November 2015}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The model can be viewed as a straightforward combination of the
maximum-entropy method and basic sampling relations from the
probability-calculus, discussed in \sect~\ref{sec:model}. The
maximum-entropy or minimum-relative-entropy method
\citep{jaynes1957}[\chap~11]{jaynes1994_r2003}[for the minimum-relative
entropy method see:][\sect~4.b]{jaynes1963}[for a more rigorous
derivation]{hobsonetal1973}[for applications]{sivia1996_r2006}
%[see also][for applications to irreversible statistical mechanics]{grandy1980}
to build initial probability distributions has been used in the
neurosciences for several kinds of investigations about the neuronal
activity of various brain areas and about other phenomena of importance % ,
% for example gene and protein interaction
\citep[]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,roudietal2009c,barreiroetal2010,shimazakietal2015,moraetal2015%,lezonetal2006,weigtetal2009
}[for
further references see][]{lathametal2013}. However, our model can also be
viewed as an approximation of a more detailed model within Bayesian theory;
we shall constantly keep this point of view in mind in our discussion.

@@

% Before applying the formula above to two concrete data sets we want to add
% two remarks about the maximum-entropy method that are seldom made in the
% neuroscientific literature. They are important for the interpretation of
% the results and are further discussed in \sect~\ref{sec:assumptions}.
% First, the maximum-entropy
% method % (based on the \emph{Shannon} relative
% % entropy, as opposed to other entropy functions such as Burg's
% % \citeyear{burg1975})
% rests on some implicit assumptions about the probabilities for the long-run
% frequency distribution of activities, besides the assumptions just
% mentioned at the end of the previous paragraph. So the often-heard
% statement that it gives \enquote{the maximally unbiased (or non-committal)
%   distribution} must be taken with a grain of salt.
% \citep{jaynes1986d_r1996,portamana2009,portamana2017}. Second,
A maximum-entropy distribution like $\px(\yA \| M, N)$ is equivalent to the
zeroth-order Laplace approximation
\citep[\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four distinct distributions for the larger population, which differ
numerically from one another in higher-order approximations:
\begin{enumerate}[label=(\roman*)]
\item\label{item:prob_freq} the most probable \emph{frequency} distribution
  for the total activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item\label{item:prob_binin} the \emph{belief} distribution for the value
  of the total activity at any time bin among the \emph{recorded} ones,
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item\label{item:prob_longrun} the most probable \emph{frequency}
  distribution for the total activity in a very long run of \emph{new} time
  bins, %$\ynuu(\yA)$,
\item\label{item:prob_binout} the \emph{belief} distribution for the value
  of the total activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{enumerate}
Their mathematical relationship is presented more explicitly in
appendix~\ref{sec:inference_prob}. The
distribution~\eqref{eq:app_maxent_pop_first} can therefore be interpreted
in each of the four ways above, in this approximate sense. In the present
case the validity of the approximation decreases as the ratio $nN/T$
increases. The most accurate interpretation is~\ref{item:prob_freq}, and that
is how we prefer to interpret the distribution $\px$: as \emph{the most
  probable relative-frequency distribution} of the activity $\yA$. See
\sect~\ref{sec:assumptions} for further discussion about assumptions and
approximations.

@@

Our question is whether some degree of \enquote{cooperation} is more
\enquote{important} than some other degree, or not. In terms of
informational sufficiency, its definition might suggest to compare the
belief distributions conditional on two different subsets of statistics.
But which measure should be used for such a comparison? and how to motivate
it? We prefer to take a different approach, leading to an automatic answer:
let us turn the informational sufficiencies of two different subsets of
statistics into two \emph{hypotheses}. We can then ask: which hypothesis is
more probable, given the data? Equivalently, what is the ratio of their
probabilities? By translating our question about \enquote{cooperativity}
into a hypothesis-testing problem about sufficiency, we can find a
well-defined answer determined by the probability-calculus. % The ratio of
% the two probabilities gives us a natural measure of the
% For example, do
% the data say that the first two empirical moments are probably enough to
% determine our beliefs? or is it more probable that the first four are
% needed instead?
We shall see that this approach is indeed more powerful than a simple
comparison of distributions.

@@

The second assumption, typical of maximum-entropy in general, concerns the
probability distribution for the long-run frequencies ensuing from the
first assumption \citep[and references therein]{portamana2017}. It is taken
to be either an \enquote{entropic prior}
\citep{skillingetal1984,rodriguez1991,neumann2007}, which takes into
account the multiplicity of long-run frequencies, or a prior for a model
with sufficient statistics, which expresses the sufficiency of the moments
under study, according to the theorem discussed in
\sect~\ref{sec:marginalization}. This assumption is crucial for the
maximum-entropy method based on the Shannon entropy. Alternative prior
densities lead to alternative maximum-entropy methods; for example one
based on the Burg entropy \citep{burg1975} if a Dirichlet density is used
\citep{jaynes1986d_r1996}. The entropic prior depends on a reference
distribution and on a coefficient $L$ which determines the sharpness of the
prior's peak. The computations of \sects~\ref{sec:application}
and~\ref{sec:marginalization} were insensitive to the choice of the
reference distribution in their significant digits. We used three
biologically reasonable alternatives: a flat distribution, one linearly
decreasing with $\yA$, and a normal one with a broad peak at low
activities.

@@

The two infinity approximations are at the boundary of their validity in
our examples. For this reason we interpret the maximum-entropy distribution
as the \emph{most probable frequency distributions}, as mentioned at the
end of \sect~\ref{sec:model}. An important question, then, is \emph{which
  features of the maximum-entropy frequency distribution are typical of the
  majority of plausible frequency distributions?} Especially for features
such as number, position, height of modes, or tail behaviour. Preliminary
studies using a full probability calculation (to be published separately)
show that several semi-quantitative features of the maximum-entropy
distribution are indeed typical. For the first data set, for example, most
frequency distributions have two high-frequency regimes of activity, as in
\fig~\ref{fig:plots_5mom_a}, with roughly the same heights and almost the
same locations, within a few percentages of population-averaged activity.
Each of the two high-frequency regimes, though, may consist of two or three
very close modes rather than one. The same preliminary studies also
indicate that these typical features are insensitive to alternative prior choices
in the second assumption above, for example using a Dirichlet prior instead
of an entropic one.
